% Suban las voces, setting II, Ambiela, estribillo
\begin{poemtitleblock}
    \poemhead{\wtitle{Suban las voces al cielo} (Lleida, before 1689)}
    \poemsource{%
        Anonymous, from musical setting by Miguel Ambiela
        (\signature{E-Bbc}{M733/1})%
    }
\end{poemtitleblock}

\begin{poemtranslation}
    \begin{original}
        \StanzaSection{4}[\add{Estribillo}]
        Suban las voces al cielo &
        y diga Capilla Regia &
        que María al cielo parte &
        pero nunca más entera. \&

        \Stanza{10}
        Y digan mudando el aire &
        en veloces corcheas, &
        vuelen, vuelen juntos &
        en síncopas que elevan, &
        con bemoles blandos:  &
        trinados que suspendan &
        sigan sus pasos todos &
        hasta la esfera  &
        donde María goze  &
        glorias eternas.
        \SectionBreak

        \StanzaSection{6}[Coplas]
        María se parte al cielo, &
        donde tendrá gloria eterna, &
        y se me parte el alma &
        de pensarse me aleja. &
        Y así el alma con voces & 
        digan en tanta pena: \&

        \Stanza{4}
        \emph{Suban las voces al cielo} &
        \emph{y diga Capilla Regia} &
        \emph{que María al cielo parte} &
        \emph{pero nunca más entera.}
        \SectionBreak

        \Stanza{6}
        Deja el mundo miserable, &
        que el mundo todo es miseria, &
        en donde liberal &
        mostrarse al mundo pueda. &
        Y así pues esperamos &
        divinas influencias: \&
        
        \Stanza{10}
        \emph{Digan mudando el aire} &
        \emph{en veloces corcheas,} &
        \emph{vuelen, vuelen juntos} &
        \emph{en síncopas que elevan.}
        \SectionBreak

        \Stanza{6}
        Tan llena de gracias sube, &
        que estéril deja la tierra, &
        si no es que a su contacto &
        fructificada queda. &
        Y así conmigo acordes &
        digan voces excelsas: \&

        \Stanza{6}
        \emph{con bemoles blandos:}  &
        \emph{trinados que suspendan} &
        \emph{sigan sus pasos todos} &
        \emph{hasta la esfera}  &
        \emph{donde María goze}  &
        \emph{glorias eternas.} \&
    \end{original}

    \begin{translation}
        \Stanza{4}
        Let voices ascend to Heaven &
        and let the Royal Chapel say &
        that to Heaven Mary departs, &
        though more whole than ever. \&

        \Stanza{10}
        And let them say, transforming the air &
        into rapid quavers, &
        let them fly, fly together &
        in syncopations that ascend &
        with mild flats: &
        trills that suspend &
        let everyone follow her paces &
        up to the sphere &
        where Mary enjoys &
        eternal glories. \&

        \StanzaSection{6}
        Mary departs for Heaven &
        where she will have glory eternal, &
        and my soul departs from me, &
        thinking on this, it leaves me far behind. &
        And thus, let the soul loudly/with voices &
        say in such sorrow: \&

        \Stanza{4}
        \emph{Let voices ascend to Heaven} &
        \emph{and let the Royal Chapel say} &
        \emph{that to Heaven Mary departs,} &
        \emph{though more whole than ever.}
        \SectionBreak

        \Stanza{6}
        She leaves the world miserable, &
        for all the world is misery &
        where she can show herself &
        to be free from the world and merciful to it. &
        And thus we await &
        divine influences: \& % XXX [from her intercession]

        \Stanza{4}
        \emph{Let them say, transforming the air} &
        \emph{into rapid quavers,} &
        \emph{let them fly, fly together} &
        \emph{in syncopations that ascend.}
        \SectionBreak

        \Stanza{6}
        She ascends so full of grace &
        that she leaves the earth sterile &
        if it were not that at her touch &
        it remains fruitful. &
        Y and thus in harmony with me &
        let the loftiest voices say: \&

        \Stanza{6}
        \emph{With mild flats,} &
        \emph{trills that suspend,} &
        \emph{let everyone follow her paces} &
        \emph{up to the sphere} &
        \emph{where Mary enjoys} &
        \emph{eternal glories.} \&
    \end{translation}
\end{poemtranslation}
\endinput

